Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-29 cis-perfect-testing.tex
More on new version
proof

I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
To be considered, an individual must have had at least three tests before time $T$ (a negative prior to the episode, a positive, and a negative following the episode).
I index these individuals by $i = 1, \dots, \Ncis$.

@@

I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}).
I assume the prior $\ntot \dist \NBc(\mu, r)$, independently of \vec{\theta}, the parameters of the survival distribution.
Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.

@@

Step 3 requires a distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
I modify the tail based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
Walker uses survival analysis to estimate the duration from CIS data.
This analysis assumes the initiating event is known, and equal to the episode’s detection time. 
It assumes the final event is interval censored between the time of the final positive test and the subsequent negative test, or right censored if a negative test has not yet been observed.
A flexible, spline-based form is used for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. \label{perf-test:fig:duration-dist}}
\end{figure}

>>>>

I consider the cohort of $\Ncis = \numprint{437590}$ individuals who could have had a detected episode; this is a subset of the whole CIS cohort.
To be considered, an individual must have at least one test between time 1 and $T$.
I index these individuals by $i = 1, \dots, \Ncis$.

@@

I derive an analytical solution to $\eta$ (defined in \cref{perf-test:eq:eta}).
I assume the prior $\ntot \dist \NBc(\mu, r)$ is independent of $\vec{\theta}$, the parameters of the survival distribution.
Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.

@@

Step 3 requires a distribution of the duration of detectability.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to be consistent with the CIS.
I modify the tail based on an unpublished analysis of CIS data~\citePersonalComms{A.\ Sarah Walker}.
Walker uses survival analysis to estimate the duration from CIS data.
This analysis assumes the initiating event is known, and equal to the episode’s detection time. 
It assumes the final event is interval censored between the time of the final positive test and the subsequent negative test, or right censored if a negative test has not yet been observed.
A flexible, spline-based form is used for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimate.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. \label{perf-test:fig:duration-dist}}
\end{figure}

### 2024-03-29 conclusion.tex
proof ch 8
proof

Pandemics will continue to pose a significant threat to public health for the foreseeable future.
Therefore, there is a public health need to respond to pandemics.
Such a response needs to be informed by monitoring.
In this thesis, I have investigated one method of monitoring pandemics: prevalence surveys.

@@

In \cref{E-intro:sec:aims}, I posed the central question for this thesis: can prevalence surveys alone be used to infer incidence and transmission, and if so how?

In \cref{E-transmission}, I found that this it is possible, developing two models that can infer incidence and transmission from prevalence surveys.
One model was phenomenological in nature and the other mechanistic.
The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
Both sets of results highlighted the heterogeneity across regions and age groups in England.
The choice of modelling approach should depend on the situation.
The phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
However, the mechanistic approach provides more information.
Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.

The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}.
This work developed the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
The major insight was the justification of the deterministic model framework and the importance of the distribution of the duration of positivity.

These models both required knowledge of the distribution of the duration of positivity.
Estimating this distribution was the focus of most of this thesis.
In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated the CIS prevalence survey.
This did require external information from the ATACCC study (the results of \cref{E-ATACCC}), demonstrating that the CIS's design is not sufficient for estimating transmission.
However, future surveys may be able to avoid this issue (see \cref{conclusion:sec:future-work}).

Prior to this thesis, the duration of positivity in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
A particularly understudied area was quantifying the shape and size of the tail of the distribution, due to limited follow-up of individuals who test positive.

Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
A well-designed survey is sufficient to estimate all the quantities of interest (as described in \cref{E-intro:sec:metrics}), except severity.

@@

@@

@@

The challenges in estimating the distribution of the duration of positivity originated in the design of the studies used.
Exploring the design of future studies to better estimate the distribution of the duration of positivity would be beneficial.
A particularly promising direction would be intensive follow-up of individuals following a positive test (\eg daily testing) until they have two or more negative tests.
Intensive follow-up would increase the precision by which the end of the infection episode is known.
Furthermore, it reduces the impact of false negatives because they are likely to be followed by a true positive.
I have developed software that could be used to perform simulation studies of alternative designs.

Other design questions are also posed by the work in this thesis.
For example, whether a representative sample of the population is optimal, or if certain groups should be oversampled.
Certain groups, such as younger individuals, were more likely to be infected.
It is plausible that sampling in proportion to incidence would be beneficial.
This is especially true if policy questions (\eg closing of schools) are of importance.
These questions could be investigated through simulation studies.
In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of positivity).

>>>>

Pandemics will continue to pose a significant threat to public health for the foreseeable future.
Therefore, there is a need for a public health response to pandemics, informed by surveillance.
In this thesis, I have investigated the potential of prevalence surveys, using the CIS as an example, to provide this information.

@@

In \cref{E-intro:sec:aims}, I posed the central question for this thesis: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?

In \cref{E-transmission}, I found that this it is possible by developing two models that can infer incidence and transmission from prevalence surveys.
These did both rely on knowledge of the distribution of the duration of detectability, which can partially be inferred from the CIS.
One model was phenomenological in nature and the other mechanistic.
The phenomenological approach, based on the backcalculation framework, provides estimates of incidence and makes fewer assumptions about the infection process; however, the estimates are very uncertain.
The mechanistic approach, based on an SEIR-type compartmental model, reduces the uncertainty and allows estimation of the components of transmission by making stronger assumptions about the infection process.
The choice of modelling approach should depend on the situation.
A phenomenological approach is more robust when there are high levels of uncertainty; for example, early in an outbreak or in the presence of new variants the infectious period duration may not be known.
On the other hand, a mechanistic approach provides more information.
Where possible, both methods should be implemented; discrepancies between results may highlight deficiencies in modelling assumptions.

The work in \cref{E-transmission} relied on the theoretical work of \cref{E-inc-prev}, adapting the backcalculation framework for the context of prevalence surveys for SARS-CoV-2.
The major insight was the justification of a deterministic backcalculation and the importance of the distribution of the duration of detectability.

Estimation of incidence and/or transmission required knowledge of the distribution of the duration of detectability.
Estimating this distribution was the focus of most of this thesis.
In \cref{E-perf-test,E-imperf-test}, I showed how this quantity could be estimated from the CIS prevalence survey.
This did require external information from the ATACCC study and prior knowledge on the number of infections, demonstrating that some additional studies are required in addition to the CIS, at least with its current design.

Prior to this thesis, the duration of detectability in the general population was not well understood (see \cref{E-intro:sec:previous-duration-estimates}).
A particularly understudied area was quantifying the shape and size of the tail of the distribution, due to limited follow-up of individuals who test positive.

Overall, I have found that representative prevalence surveys are a promising approach for pandemic surveillance.
A well-designed survey is sufficient to estimate incidence and transmission, although not currently the duration of detectability.

@@

\subsection{Joint modelling}

@@

\subsection{Study design}

@@

The challenges in estimating the distribution of the duration of detectability originated in the design of the studies used.
Exploring the design of future studies to better estimate the distribution of the duration of detectability would be beneficial.
A particularly promising direction would be intensive follow-up of individuals following a positive test (\eg daily testing) until they have two or more negative tests.
Intensive follow-up would increase the precision by which the end of the infection episode is known.
Furthermore, it reduces the impact of false negatives because they are likely to be followed by a true positive.
I have developed software that could be used to perform simulation studies of alternative designs.

Other design questions are also posed by the work in this thesis.
For example, whether a representative sample of the population is optimal, or if certain groups should be oversampled.
Certain groups, such as younger individuals, were more likely to be infected.
It is plausible that sampling in proportion to incidence would be beneficial.
This is especially true if policy questions such as closing of schools are of importance.
These questions could be investigated through simulation studies.
In constructing such studies, it will be important to consider how these considerations might be different for different pathogens (\eg influenza, the most common source of respiratory pandemics) and tests (\eg rapid antigen tests, which have a shorter duration of detectability).

\subsection{Other extensions}

### 2024-03-29 introduction.tex
proof

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of positivity.
However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of positivity.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of positivity.
However, the study's cohort consisted only of healthcare workers, \ie working-age adults, and included only 27 infections.
Therefore, the data are not suitable for inferring the full distribution in the general population.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
Furthermore, asymptomatic infections were excluded due to limitations of their methodology.
Duration of positivity may be correlated with severity of symptoms, and hence this exclusion could bias their results.

\Textcite{ealesCharacterising} estimated the duration of positivity from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of positivity).
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of positivity.

For the purposes of this thesis, I require estimates applicable to the general population.
There are no reliable estimates of the full distribution of the duration of positivity in a general population cohort available in the literature.
Therefore, I estimate the duration of positivity from the ATACCC and CIS studies.

@@

Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
% These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
% The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
Their model structure limits the form of the duration of positivity. 
They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
% These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
They allow an arbitrary distribution of the duration of positivity, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of positivity and the susceptibility of the population.
Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.
The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of positivity from a general population study, unlike prior work.

@@

Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
I allow a more flexible distribution of the duration of positivity than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
I use only data from the CIS and show this is sufficient for estimating incidence and transmission.

@@

First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of positivity.
This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently individuals should be tested, or how many individuals should be tested.

Second, the survival analysis framework used to infer the duration of positivity in \cref{E-perf-test,E-imperf-test}.
The framework allows the user to specify all the different analyses performed in these chapters in a simple interface. 
Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.

@@

In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
A key insight in this chapter is that the distribution of the duration of positivity is the key quantity linking incidence and prevalence.

I then turn to estimating this distribution.
I start, in \cref{E-ATACCC}, by estimating the duration of positivity from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
Estimating the tail requires a large sample size and long follow-up. 
Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
In \cref{E-perf-test}, I provide a framework for estimating the duration of positivity from the CIS data, assuming perfect tests.
However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.

Having estimated the duration of positivity, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
Finally, I conclude in \cref{E-conclusion}.

>>>>

The second study is the CIS (introduced in \cref{intro:sec:data-for-estimating}).
In addition to its role as a prevalence survey, the CIS followed the same individuals over a long period of time (over a year in many cases), collecting longitudinal data.
The large sample size and long follow-up allows estimation of the tails of the distribution of the duration of detectability.
However, the lack of frequent testing (at most weekly) prohibits estimation of the viral load dynamics.
In particular, the vast majority of detected infections had only one or two positive tests, and hence the viral load trajectory of these infections cannot be estimated.
I describe the CIS in detail in \cref{E-intro:sec:cis} and use it to estimate duration in \cref{E-perf-test,E-imperf-test}.

@@

An early meta-analysis estimated that individuals test positive for SARS-CoV-2 for a mean of 14.6 days (95\% CI: 9.3--20.0 days)~\autocite{cevikShedding}.
The longest reported duration was observed at 83 days, although beyond day 9 no live virus was detected, indicating that the individuals testing positive for longer than this were not infectious.
However, most studies included in the meta-analysis involved hospitalised patients or had unclear inclusion criteria, and hence may not be representative of the general population.
These summary estimates are insufficient for backcalculation, which requires a distribution of the duration of detectability.

\Textcite{hellewellPCRSensitivity} estimated the probability of an individual testing positive each day after infection.
This measure is highly related to the duration of detectability.
However, the study's cohort consisted only of healthcare workers, \ie working-age adults, and included only 27 infections.
Therefore, the data are not suitable for inferring the full distribution in the general population.

\Textcite{binnySensitivity} applied a slight variant of \textcite{hellewellPCRSensitivity}'s methodology to data from New Zealand.
The situation in New Zealand, who were successfully pursuing a zero-COVID strategy at this time, means that the individuals in the study were unusual.
For example, a third were travellers detected during their quarantine period on arrival to New Zealand but would have had negative RT-PCR tests before their departure.
Furthermore, asymptomatic infections were excluded due to limitations of their methodology.
Duration of detectability may be correlated with severity of symptoms, and hence this exclusion could bias their results.

\Textcite{ealesCharacterising} estimated the duration of detectability from REACT data.
Specifically, individuals who tested positive in round 8 of REACT (6th--22 Jan 2021) were invited to take two additional swab tests.
Defining the time of the initial positive test at time 0, \textcite{ealesCharacterising} estimated the probability of remaining positive at each day after the initial positive test.
They made a strong parametric assumption that the probability decreases exponentially, with a sensitivity analysis assuming a plateau before the decay.
This assumption is necessary because they had at most 17 days of follow-up, and hence could not estimate the tail of the distribution.
Overall, they estimate a median duration of 14.0 days (95\% CI: 12.9--15.4 days) with substantial positive correlation with each individual’s viral load at the time of the initial positive test (\ie a higher initial viral load was associated with a longer duration of detectability).
The individuals with the highest viral load individuals were estimated to have a median duration of 19.6 days (16.1--25.8 days).
The association could be due to individuals with a higher initial viral load being (on average) more recently infected at the time of the initial positive test.
The study design prohibits estimating how long before the initial positive test the infection occurred.
The same issue means the study underestimates the duration of detectability.

For the purposes of this thesis, I require estimates applicable to the general population.
There are no reliable estimates of the full distribution of the duration of detectability in a general population cohort available in the literature.
Therefore, I estimate the duration of detectability from the ATACCC and CIS studies.

@@

Several compartmental models incorporated data from SARS-CoV-2 prevalence surveys, alongside other data sources~\autocite{daviesAssociation,ironsEstimating,knockKey,nicholsonImproving,pooleyEstimation,birrellRTM2}.
% These models incorporate prevalence survey data by adding additional compartments representing RT-PCR positive individuals, who may not be infectious (this is common, see \cref{E-biology-data}); except \textcite{nicholsonImproving} who have a more sophisticated approach (below) and \textcite{ironsEstimating}, who assume RT-PCR positivity coincides with infectiousness.
% The length of stay in these compartments (the duration of time positive) is assumed to follow a gamma distribution, with the parameters estimated from external data or having at most one free parameter.
Their model structure limits the form of the duration of detectability. 
They fit the models to prevalence survey data using a binomial likelihood where possible, although some approaches take a normal approximation.
% These approaches are similar to how these models are fit to other data sources, such as hospital admissions or deaths.

\Textcite{nicholsonImproving} fit a simple mechanistic model, notably without age structure, but have the most sophisticated observation model.
They allow an arbitrary distribution of the duration of detectability, based on the estimates of \textcite{hellewellPCRSensitivity} (see \cref{intro:sec:previous-duration-estimates} for discussion of the \textcite{hellewellPCRSensitivity} study).
Furthermore, they incorporate the sensitivity and specificity of the RT-PCR test into their model.

@@

In order to achieve these aims, I use a case study, specifically the SARS-CoV-2 epidemic in England from Mon 31 Aug 2020 until Sun 25 Jan 2021.
This case study has several features that make it useful for the aims of this thesis.
These include that the CIS was of sufficient size and incidence sufficiently high that there is enough data to perform the analyses.
Additionally, incidence begins rising over this period, which is the period when a surveillance system needs to provide a warning of upcoming issues.
The CIS began in April 2020, when incidence was falling due to the first lockdown in the UK.
Incidence remained low and fairly flat in England over summer 2020, until it began rising in mid or late August.
Finally, the period ends before the roll-out of the vaccine, which likely affects many important aspects of the analysis.
In particular, the duration of detectability and the susceptibility of the population.
Including both of these elements would require more complex modelling and including idiosyncratic features of the SARS-CoV-2 pandemic that are less generalisable to other diseases.

@@

First, in \cref{E-inc-prev:sec:observation-process}, I provide a theoretic justification for the use of deterministic backcalculation in this setting.
A \emph{deterministic} backcalculation approach was used by \textcite{abbottCISincidence} and I will use it in \cref{E-backcalc}.
The deterministic approach models the prevalence at any time as a deterministic function of the incidence, rather than as a random variable.
I show that this is a good approximation as long as the proportion of the population sampled is small.
In particular, it does not require the prevalence or incidence to be large.
The estimates I produce in \cref{E-backcalc} fully propagate the uncertainty from the prevalence estimation procedure and use estimates of the duration of detectability from a general population study, unlike prior work.

@@

Finally, in \cref{E-SEIR}, I describe a compartmental model for SARS-CoV-2 transmission, which I fit to CIS data.
I allow a more flexible distribution of the duration of detectability than any of the papers discussed above, except \textcite{nicholsonImproving}, incorporating the skew of the distribution.
This distribution is based on data from a representative sample of the UK population, rather than convenience samples (see \cref{intro:sec:previous-duration-estimates}).
Unlike \textcite{nicholsonImproving}, I include age-stratification in the transmission model, allowing estimation of how transmission varies by age.
I use only data from the CIS and show this is sufficient for estimating incidence and transmission.

@@

First, the framework to simulate the CIS (used in \cref{E-imperf-test:sec:simulate,E-perf-test:sec:simulation-study}).
The package enables customisation of various features of the study, for example the testing regime of the participants, the sensitivity of the tests, and the duration of detectability.
This package would be particularly useful in answering questions regarding the design of future prevalence surveys, for example, how frequently individuals should be tested, or how many individuals should be tested.

Second, the survival analysis framework used to infer the duration of detectability in \cref{E-perf-test,E-imperf-test}.
The framework allows the user to specify all the different analyses performed in these chapters in a simple interface. 
Furthermore, adding new analyses, for example varying the priors or the inclusion of covariates, is straightforward.

@@

In \cref{E-biology-data}, I provide background information on the natural history of SARS-CoV-2, the RT-PCR testing process used to produce the data used throughout this thesis, and describe the two cohort studies used in this thesis.
Then, in \cref{E-inc-prev}, I formally describe the stochastic processes which generate infections, prevalence, and the observations of prevalence.
A key insight in this chapter is that the distribution of the duration of detectability is the key quantity linking incidence and prevalence.

I then turn to estimating this distribution.
I start, in \cref{E-ATACCC}, by estimating the duration of detectability from the ATACCC study.
This provides a reliable estimate of the bulk of the distribution (out to around 20 days) but shows that there is a significant tail.
Estimating the tail requires a large sample size and long follow-up. 
Therefore, I turn to the CIS data in \cref{E-perf-test,E-imperf-test}.
In \cref{E-perf-test}, I provide a framework for estimating the duration of detectability from the CIS data, assuming perfect tests.
However, the assumption of perfect tests is not realistic; in \cref{E-imperf-test}, I extend the framework to include false negatives.

Having estimated the duration of detectability, next I estimate incidence and transmission from the CIS data in \cref{E-transmission}, thus fulfilling the primary aim of this thesis.
I use both a phenomenological and mechanistic transmission model, contrasting their results and discussing the advantages and disadvantages of each in this context.
Finally, I conclude in \cref{E-conclusion}.

### 2024-03-29 introduction_new.tex
proof

The overall question that I investigate in this thesis is whether a prevalence survey \emph{alone} is sufficient to infer incidence and transmission and, if so, what methodology can be used.
% First, how can prevalence surveys be designed to help estimate these quantities?

>>>>

The overall question that I investigate in this thesis is: can a prevalence survey \emph{alone} be sufficient to infer incidence and transmission and, if so, what method(s) are optimal?
% First, how can prevalence surveys be designed to help estimate these quantities?

### 2024-03-29 paper-surv-analysis.tex
proof

For privacy reasons, the CIS data is stored in the ONS's SRS (Secure Research Service).
The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
The SRS must be used to perform analyses using individual-level data.
Therefore, computationally efficient approaches must be taken to this work.

@@

In forming the probability $p_{ia}$, the data is used.
This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

@@

To generate the durations, an assumption is required for the distribution of the duration of positivity.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on a previously unpublished analysis of CIS data, without accounting for several of the issues previously described.
Walker uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. See main text for details of each analysis. \label{perf-test:fig:duration-dist}}
\end{figure}

>>>>

For privacy reasons, the CIS data are stored in the ONS's SRS (Secure Research Service).
The SRS is a Trusted Research Environment (TRE) that allows researchers to access sensitive data in a secure environment.
The SRS must be used to perform analyses using individual-level data.
Therefore, computationally efficient approaches must be taken to this work.

@@

In forming the probability $p_{ia}$, the data are used.
This is not quite correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own class is possible.
Then, the counts for all the classes that previously contained undetected episodes would be zero.
The resulting likelihood would be the same.
Therefore, I present this version as it is clearer.

@@

To generate the durations, an assumption is required for the distribution of the duration of detectability.
I base this assumption on the estimate from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on a previously unpublished analysis of CIS data, without accounting for several of the issues previously described.
Walker uses survival analysis to estimate the duration from CIS data.
The initiating event is assumed known as the time the episode was detected.
The final event is assumed interval censored between the time of the final positive test and the subsequent negative test, or right-censored if a negative test has not yet been observed.
A flexible, spline-based form is assumed for the baseline survival function~\autocite{roystonSTPM,roystonFlexible} with covariates introduced via proportional odds.
By not accounting for either the undetected infections or the interval censoring of the initiating event, this analysis has competing biases which makes them hard to interpret~\autocite{cisMethodsONS}.

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The first 30 days of the distribution is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of detectability, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. See main text for details of each analysis. \label{perf-test:fig:duration-dist}}
\end{figure}
